Supplementary Material

### 2022-11-09 paper/paper/SPIE/supplementary.tex
make supplementary material

\documentclass[12pt]{spieman}  % 12pt font required by SPIE;
%\documentclass[a4paper,12pt]{spieman}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{graphicx}
\graphicspath{ {./figures/} }
\usepackage{setspace}
\usepackage{tocloft}
\usepackage{doi}
\usepackage{csquotes}
\MakeOuterQuote{"}
\usepackage{xr}
\externaldocument[S-]{supplementary}

@@

\title{Random Matrix Theory Tools for the Analysis of Functional Magnetic Resonance Imaging Examinations}

@@

\renewcommand{\cftdotsep}{\cftnodots}
\cftpagenumbersoff{figure}
\cftpagenumbersoff{table}
\begin{document}
\maketitle

% \begin{abstract} This document shows the required format and appearance of a
% manuscript prepared for SPIE journals. It is prepared using LaTeX2e with the
% class file \texttt{spieman.cls}. Please note that the following journals
% require the use of structured abstracts in manuscript submissions:
% \textit{Neurophotonics}, the \textit{Journal of Biomedical Optics}, and the
% \textit{Journal of Medical Imaging}. Structured abstracts are encouraged for
% the \textit{Journal of Micro/Nanolithography, MEMS, and MOEMS}. Guidelines
% are available on the journal website. Whether structured or single-paragraph,
% the abstract should be a summary of the paper and not an introduction.
% Because the abstract may be used in abstracting and indexing databases, it
% should be self-contained (i.e., no numerical references) and substantive in
% nature, presenting concisely the objectives, methodology used, results
% obtained, and their significance. A list of up to six keywords should
% immediately follow. \end{abstract}

% See
% https://www.spiedigitallibrary.org/journals/journal-of-medical-imaging/volume-7/issue-01/010101/Integrating-structured-abstracts-in-the-Journal-of-Medical-Imaging/10.1117/1.JMI.7.1.010101.full
% for info on the structured abstract
%
% Purpose: This section presents the significance and aims, stating the broad
% impact and the rationale or motivation of the work, including potentially
% some background.
%
% Approach: This section briefly describes the materials and methods, including
% the statistical analyses, used in the research.
%
% Results: This section provides a core summary of the findings from the work,
% including study numbers, quantitative analyses, or discoveries.
%
% Conclusions: This section summarizes and interprets the approach and findings
% from the work, stating also the importance or impact of the findings.

% Purpose: Data-intensive modeling could provide insight on the broad
% variability in outcomes in spine surgery. Previous studies were limited to
% analysis of demographic and clinical characteristics. We report an analytic
% framework called “SpineCloud” that incorporates quantitative features
% extracted from perioperative images to predict spine surgery outcome.

% Approach: A retrospective study was conducted in which patient demographics,
% imaging, and outcome data were collected. Image features were automatically
% computed from perioperative CT. Postoperative 3- and 12-month functional and
% pain outcomes were analyzed in terms of improvement relative to the
% preoperative state. A boosted decision tree classifier was trained to predict
% outcome using demographic and image features as predictor variables.
% Predictions were computed based on SpineCloud and conventional demographic
% models, and features associated with poor outcome were identified from
% weighting terms evident in the boosted tree.

% Results Neither approach was predictive of 3- or 12-month outcomes based on
% preoperative data alone in the current, preliminary study. However,
% SpineCloud predictions incorporating image features obtained during and
% immediately following surgery (i.e., intraoperative and immediate
% postoperative images) exhibited significant improvement in area under the
% receiver operating characteristic (AUC): AUC=0.72 (CI95=0.59 to 0.83) at 3
% months and AUC=0.69 (CI95=0.55 to 0.82) at 12 months.

% Conclusions Predictive modeling of lumbar spine surgery outcomes was improved
% by incorporation of image-based features compared to analysis based on
% conventional demographic data. The SpineCloud framework could improve
% understanding of factors underlying outcome variability and warrants further
% investigation and validation in a larger patient cohort.
\begin{abstract}

\section*{Purpose}
Random matrix theory (RMT) is an increasingly useful tool for understanding
large, complex systems. Prior studies have examined functional Magnetic
Resonance Imaging (fMRI) scans using tools from RMT, with some success.
However, RMT computations are highly sensitive to a number of analytic choices,
and the robustness of findings involving RMT remain in question. We
systematically investigate the usefulness of RMT on a wide variety of fMRI
datasets using a rigorous predictive framework.

\section*{Approach}
We develop open-source software to efficiently compute RMT features from fMRI
images, and examine the cross-validated predictive potential of eigenvalue and
RMT-based features (``eigenfeatures'') with classic machine-learning
classifiers. We systematically vary pre-processing extent, normalization
procedures, RMT unfolding procedures, and feature selection, and compare the
impact of these analytic choices on the distributions of cross-validated
prediction performance for each combination of dataset binary classification
task, classifier, and feature. To deal with class imbalance, we use the
area under the receiver operating characteristic (AUROC) as the main
performance metric.

\section*{Results}
% df.groupby(["subgroup", "gross_feature"])["auroc"].median()
Across all classification tasks and analytic choices, we find eiegenfeatures to
more often have predictive utility than not (82.4\% of median AUROCs \(>\) 0.5;
median AUROC range across classification tasks 0.47 - 0.64). Simple baseline
reductions on source timeseries, by contrast, were less useful (58.8\% of
median AUROCs \(>\) 0.5, median AUROC range across classification tasks 0.42 -
0.62). Additionally, eigenfeature AUROC distributions were overall much more
right-tailed than baseline features suggesting greater predictive potential.
However, performance distributions were wide, and often significantly affected by
analytic choices.

\section*{Conclusions}
Eigenfeatures clearly have potential for understanding and reducing fMRI
functional connectivity. However, the utility of these features is strongly
dependent on analytic decisions, suggesting considerable caution when
interpreting past and future studies applying RMT to fMRI.

% This document shows the required format and appearance of a manuscript
% prepared for SPIE journals. It is prepared using LaTeX2e with the class file
% \texttt{spieman.cls}. Please note that the following journals require the use
% of structured abstracts in manuscript submissions: \textit{Neurophotonics},
% the \textit{Journal of Biomedical Optics}, and the \textit{Journal of Medical
% Imaging}. Structured abstracts are encouraged for the \textit{Journal of
% Micro/Nanolithography, MEMS, and MOEMS}. Guidelines are available on the
% journal website. Whether structured or single-paragraph, the abstract should
% be a summary of the paper and not an introduction. Because the abstract may
% be used in abstracting and indexing databases, it should be self-contained
% (i.e., no numerical references) and substantive in nature, presenting
% concisely the objectives, methodology used, results obtained, and their
% significance. A list of up to six keywords should immediately follow.
\end{abstract}

% Include a list of up to six keywords after the abstract
\keywords{random matrix, spectral rigidity, level number variance, fMRI, classification, machine-learning}

% Include email contact information for corresponding author
{\noindent \footnotesize\textbf{*}Jacob Levman,  \linkable{jlevman@stfx.ca} }

\begin{spacing}{2}   % use double spacing for rest of manuscript

\section{Introduction}
\label{sect:intro}  % \label{} allows reference to this section
% This document shows the format and appearance of a manuscript prepared for
% submission to an SPIE journal. Note that this template is only intended to be
% used as a guideline for author convenience. It is designed for optimum
% clarity and ease of reading for editors and reviewers, but the template does
% not reflect the final page layout of a published journal paper. Accepted
% papers are professionally typeset in XML according to the layout and design
% of the journal.

At its most rudimentary, RMT describes the expected behaviour of the
eigenvalues—also often called levels—of a number of classes of random
matrices\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}. A random matrix is,
simplifying somewhat, a matrix in which the entries are drawn from
independently and identically distributed (\textit{iid}) distributions. Though
first developed to describe the fluctuation of nuclei energy levels in quantum
physics \cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a},
RMT has been shown have extremely broad potential.

In small-scale physical systems, RMT universalities have been observed in
quantum chaotic systems, complex nuclei, atoms, molecules and disordered
mesoscopic systems
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004,brodyRandommatrixPhysicsSpectrum1981,beenakkerRandommatrixTheoryQuantum1997,bohigasHigherOrderCorrelationsSpectra1985,wintgenLevelStatisticsQuantized1988,pandeySkewOrthogonalPolynomialsUniversality2001},
and at larger scales, RMT has been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002} , social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids
\cite{sastrySpectralStatisticsInstantaneous2001,matharooSpectralStatisticsQuenched2009}
and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,matharooVibrationalSpectraAmorphous2005,matharooUniversalityVibrationalSpectra2008}.
Within biological systems, RMT has also been used to successfully model aspects
of  amino acid functional relationships
\cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in epileptic seizures
\cite{osorioPhasesynchronizationRandommatrixBased2011}, and in protein-protein
interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
even more recently, has provided insights into the behaviours of deep
neural networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

RMT likely has the most real-world explanatory potential when dealing with a
complex system of many (hundreds or more\cite{mehtaRandomMatrices2004})
interacting components. If such a system has a matrix representation, and the
eigenvalues of this representation have a distribution similar to one predicted
by RMT, it suggests the system is either highly random or chaotic. By contrast,
if the observed spectra deviate significantly from RMT-predicted spectra, this
suggests otherwise. A number of studies have used RMT to make such
interpretations and comparisons between
systems\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003, wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016, matharooSpontaneousBackpainAlters2020}.

\subsection{RMT and Neurobiological Signals}

In the human brain, each neuron, collection of neurons, or region of interest
(ROI) is a potentially-interacting component in a highly complex system. RMT
may have potential in describing the totality of these interactions, provided
that measurements of functioning can be obtained with sufficient spatial and
temporal resolution to speak to some neurobiological or neuropsychological
phenomenon of interest.

For an imaging modality like fMRI, where changes in the
blood-oxygenation-level-dependent (BOLD) signals are related to neural
activity, RMT may be an ideal starting point. Each voxel time course (or
collection of such time sources - i.e. ROIs) can be considered an interacting
component of the system. Likewise, in \textit{functional connectivity}
analyses, statistical relationships of the BOLD ROI time-courses are
investigated in the hope of gaining insights into brain functioniong (see
\citenum{bastosTutorialReviewFunctional2016,
vandenheuvelExploringBrainNetwork2010} for reviews, but also
\citenum{nobleDecadeTestretestReliability2019} and
\citenum{lurieQuestionsControversiesStudy2020} for challenges facing functional
connectivity analyses). In this framework, each connection or correlation can
be considered a system component, and the eigenvalues of such a correlation
matrix can be examined from the perspective of RMT.

The earliest study taking this approach
demonstrated that spectra of the correlations between electroencephalographic
(EEG) signals closely resemble those of the Gaussian Orthogonal Ensemble (GOE)
\cite{sebaRandomMatrixAnalysis2003}. In fMRI, RMT has been used to
evaluate the quality of whole brain features extracted from fMRI data
\cite{voultsidouFeatureEvaluationFMRI2007,verganiRestingStateFMRI2019}, and in
diffusion MRI to aid in the selection of the number of
components to employ in principal-component reduction analysis and denoising
\cite{veraartDenoisingDiffusionMRI2016,verganiRestingStateFMRI2019,ulfarssonDimensionEstimationNoisy2008}.

RMT has also been used in ROI-based fMRI functional-connectivity studies to
investigate differences between rest and task states
\cite{wangSpectralPropertiesTemporal2015}, between subjects with and without
ADHD \cite{wangRandomMatrixTheory2016}, and between pain and non-pain states
\cite{matharooSpontaneousBackpainAlters2020}. Across these three studies, the
spectra of resting or low-attention states exhibited properties close to the
GOE. These findings suggest that certain aspects of psychological processes
might be characterized, in part, by features computed from the eigenvalues of
fMRI correlation matrices, and that these features might vary in an
interpretable manner across psychological processes. If this is the case, RMT
could aid in understanding the functioning of the human brain.

\subsection{Eigenvalues and Prediction}

The basic insight of RMT is thus that the eigenvalues alone may provide
interesting information about highly complex systems. However, real systems are
usually noisy, and involve a mixture of random and non-random components and
interactions. RMT is statistical in nature, describing only the expected
behaviour of the spectra of \textit{iid} random matrics. To this end, a number
of summary statistics or ``spectral observables''
\cite{mehtaRandomMatrices2004, guhrRandommatrixTheoriesQuantum1998a} can be
computed from empirically observed spectra, with these spectral observables
sometimes being better suited for further analysis, or for the comparison of
empirical observations to theory. Two such summary statistics that have been
popular\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014, matharooSpontaneousBackpainAlters2020,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003,wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016} are the \textit{spectral rigidity} and
\textit{level number variance} (``rigidity'' and ``level variance'' for short;
details in Section \ref{sect:methods}).

However, from a predictive standpoint, these summary statistics may eliminate
predictively-useful information. If the basic RMT insight is that the
eigenvalues alone can provide understanding of a system, then those eigenvalues
(or other simple transformations of them) ought also to be predictively
useful. We examine a number of such \textit{eigenfeatures} in this study.

In addition, the functional connectivity - the matrix of correlations of
various collections of voxels of an fMRI scan - is a priori a useful
representation of the fMRI data. However, the full correlation matrix between
all voxels is itself often computationally infeasible to work with, and thus is
reduced in various ways prior to being used in analyses.

Typical reductions include the use of a ``seed'' voxel or collection of voxels
(i.e. ROI): the mean signal over that ROI is correlated with all other N ROIs
of interest to generate a reduced functional connectivity matrix (or image)
with one correlation value at each non-seed ROI. This sort of reduction reduces
the functional connectivity to N values. This approach strongly biases the
analysis to the seed ROI.

Likewise, one can work with ROI mean signals, and work with the \(N \times N\)
matrix of these ROI correlations. This analysis [maybe cite Poldrack book for
this?] is, \textit{a priori}, sensitive to the choice of ROIs, and, in the case
of anatomical atlases / parcellations, may also lack theoretical justification.

A major disadvantage of both the above approaches, besides requiring
parcellation into ROIs, is that most such parcellation methods are atlas-based,
and thus require image registration, and, general, an entire preprocessing
pipeline to ensure that registration is successful.

RMT suggests a potentially useful reduction in the form of the eigenvalues of
the voxelwise functional connectivity. An \(N \times t\) matrix \(\bf{A}\) of
\(N\) time series of length \(t\), and where \(t \ll N\) will have a symmetric
correlation (or covariance) matrix with \(t - 1\) non-zero, positive
real-valued eigenvalues \(\Lambda\). These eigenvalues can be computed highly
efficiently via transposition of the voxelwise correlation or covariance
matrix.

\subsection{Limitations of Previous Work}

However, these previous studies took largely descriptive / explanatory
approaches, noting only differences in RMT across subgroups and/or conditions,
in many cases, with no formal statistical model nor formal testing of the
significance and robustness of these differences. No code was provided to allow
the reproduction of the results, and the extraction of the spectral rigidity
and level variance is non-trivial, and computationally demanding. Computation
of these spectral observables requires an \textit{unfolding}
procedure\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004},
which is, unfortunately, a poorly-documented fitting procedure that requires
making a number of difficult-to-justify decisions, and these decisions can
dramatically impact RMT conclusions
\cite{abul-magdUnfoldingSpectrumChaotic2014,abueleninSpectralUnfoldingChaotic2018,fossionRandommatrixSpectraTime2013,abueleninEffectUnfoldingSpectral2012,moralesImprovedUnfoldingDetrending2011}.
On top of the flexibility in the implementation and application of RMT to
empirical data, there is also the flexibility introduced by the choices in the
complex preprocessing pipelines of fMRI\cite{parkerBenefitSliceTiming2019}.

In addition these issues stemming from ``research degrees of freedom'' (cite
Gelman), there is also a basic baseline comparison issue: RMT is mathematically
non-trivial, the extraction of RMT spectral statistics is computationally
demanding, and eigenvalues, while easy to understand mathematically, are
difficult to interpret relative to simpler potential reductions of fMRI data.

What is missing is a rigorous, systematic, reproducible investigation of the
*predictice* value of RMT metrics across a wide variety of data, RMT analytic
choices, choice of eigenfeature, and choice of preprocessing, and for these RMT
eigenfeatures to be compared to simpler alternative baselines. The current
study aims to do just this.

\section{Datasets}
\label{sec:datasets}

\subsection{Overview}

We selected fMRI data publicly available on the
\href{https://openneuro.org/}{OpenNeuro} platform (36). Selection criteria was
somewhat subjective, but we required (1), that all scans have the same spatial
and temporal resolutions, (2), the the dataset comprise 10+ human subjects, and
(3), that it was reasonable to classify at the level of an entire scan, i.e.
neither task nor event timings were needed to determine class membership. This
yielded seven datasets comprising XX total subjects, and over 5000 total fMRI
volumes.

\begin{table}[h!]
\caption{
    ID = Identifier for paper. TR = Time of Repetition (seconds).
    Time = total duration (minutes) of each scan. Dimensions listed as M x N x P,
    indicate P horizontal / axial slices each with dimensions M x N.
}
\label{table:1}
\small
\centering
\begin{tabular}{ l c c c c c c }
\hline
\textbf{Name}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{n\_scans} \\
\hline
Aging     & 74 × 74 × 32   & 3.0 × 3.0 × 4.0 & 2.0  & 300 & 62  \\
Bilingual    & 100 × 100 × 72 & 1.8 × 1.8 × 1.8 & 0.88 & 823 & 90  \\
Depress   & 112 × 112 × 25 & 2.0 × 2.0 × 5.0 & 2.5  & 100 & 72  \\
Learn     & 64 × 64 × 36   & 3.0 × 3.0 × 3.0 & 2.0  & 195 & 432 \\
Osteo     & 64 × 64 × 36   & 3.4 × 3.4 × 3.0 & 2.5  & 300 & 74  \\
Park      & 80 × 80 × 43   & 3.0 × 3.0 × 3.0 & 2.4  & 149 & 552 \\
Attention & 128 × 128 × 70 & 1.5 × 1.5 × 1.5 & 3.0  & 300 & 90  \\
\hline
\end{tabular}
\end{table}

In order not to inundate readers with dataset details, and because this is an
exploratory investigation of RMT which is not committed to any specific
theory\footnote{In fact, datasets and, later, the classification tasks, were
chosen without investigation into any findings or publications associated with
the data, both in order to reduce bias in our decisions, and because the
original authors analytic / statistical approaches and intentions have very
limited relevance to our whole-brain, voxelwise, multiverse predictive
approach.}, we refer the reader to the original publications and/or data
releases when greater detail is needed, and highlight only the most basic
aspects of each dataset here.  Dataset scan parameters and acquisition details
are summarized in Table \ref{table:1}, and sample and subgroup sizes are
summarized in Table \ref{table:2}.

\begin{table}[h!]
\caption{
    ID = Identifier for paper. Subgroup = name of subgroup. Subjects = Number of subjects.
    Task = fMRI task. ANT = Attention Network Task \cite{fanActivationAttentionalNetworks2005}
}
\label{table:2}
\small
\centering
\begin{tabular}{ l r c c c }
\hline
\textbf{ID}  & \textbf{Subgroup}  & \textbf{Subjects}  & \textbf{Task} & \textbf{Scans per Subject} \\
\hline
Aging   &  older             & 28      & resting-state        & 1 \\
        &  younger           & 34      & resting-state        & 1 \\
        &                    &         &                      &   \\
Bilingual  &  bilingual          & 59      & resting-state    & 1 \\
        &  monolingual        & 33      & resting-state       & 1 \\
        &                    &         &                      &   \\
Depress &  depression         & 51      & resting-state       & 1 \\
        &  control            & 21      & resting-state       & 1 \\
        &                    &         &                      &   \\
Learn   &  task              & 24      & learn image sketches & 16\\
        &  rest              & 24      & resting-state        & 2 \\
        &                    &         &                      &   \\
Osteo   &  duloxetine        & 19      & resting-state        & 1 \\
        &  pain              & 37      & resting-state        & 1 \\
        &  nopain            & 20      & resting-state        & 1 \\
        &                    &         &                      &   \\
Park    &  Parkinson's       & 25      & ANT                  & 12\\
        &  control           & 21      & ANT                  & 12\\
        &                    &         &                      &   \\
Attention&  vigilant           & 11     & resting-state       & 2 \\
        &  nonigilant         & 11     & resting-state        & 2 \\
        &  trait-attentive     & 11    & resting-state        & 2 \\
        &  trait-non-attentive & 11    & resting-state        & 2 \\
        &  task-attentive     & 11     & resting-state        & 2 \\
        &  task-nonattentive  & 11     & resting-state        & 2 \\
\hline
\end{tabular}
\end{table}

\subsection{Datasets and Classification Tasks}

The \textit{Aging} dataset\cite{wahlheimIntrinsicFunctionalConnectivity2021}
included rs-fMRI data for 34 subjects with a mean age of 22 years (range: 18-32
years), and 28 subjects with a mean age of 70 years (range: 61-80 years). For
our analysis, we make the binary classification task the prediction of subject
age-group membership, i.e. \code{younger v older}.

% The final sample included 34 younger adults (1832 years old, Mage = 22.21, SD
% = 3.65; 20 female) and 28 older adults (61-80 years old, Mage = 69.82, SD =
% 5.64; 20 female). Table 1 shows that younger and older adults had comparable
% education and working memory capacity, the latter measured by forward and
% backward digit span (48). Older adults had higher vocabulary scores (49) and
% slower processing speed (50) than younger adults. Resting-state data
% comprised 300 measurements collected over one 10-minute run. This scan
% duration produces reliable functional connectivity estimates within and
% across participants (57). Participants were instructed to remain still and
% awake with their eyes open. No stimuli appeared during this scan, and the
% display was black. Functional scans were collected using an echo-planar image
% sequence sensitive to blood-oxygen-level-dependent (BOLD) contrast (T2*; 32
% slices with 4.0 mm thickness and no skip, TE = 30 ms, TR = 2000 ms, flip
% angle = 70, FOV = 220 mm, matrix size = 74 × 74 × 32 voxels, A/P phase
% encoding direction). Slices were collected in a descending order that covered
% the entire cortex and partial cerebellum. At the beginning of the
% resting-state scan, the scanner acquired and discarded two dummy scans.

% \subsubsection{Bilinguality Dataset}

The \textit{Bilinguality} dataset\cite{ds001747:1.0.0,
goldcarrieelizabethExploringRestingState2018} examined English and
Spanish-speaking monolinguals and multilinguals during a prolonged resting
state. The original study grouped participants into three subgroups: early vs
late bilinguals vs monolingual
controls\cite{goldcarrieelizabethExploringRestingState2018}. For our analysis,
we make the binary classification task be to classify \code{monolingual v
bilingual}.

% Resting state fMRI scans were performed using a T2*-weighted EPI sequence with the following
% parameters: TR=875ms; TE=43.6ms; slice thickness=1.8mm; acquisition matrix=100 × 100; flip
% angle=55°; number of slices=72; field of view=180 × 180mm; voxel size=1.8 × 1.8 × 1.8mm;
% multi-band factor=8; volumes per run=823 (total scan time=12min). EPI scans were oriented
% parallel to the long axis of the hippocampus.

% \subsubsection{Depression Dataset}

The \textit{Depression} dataset\cite{bezmaternykhddRestingStateClosed2021,
bezmaternykhBrainNetworksConnectivity2021} includes rs-fMRI scans frmo
non-depressed controls and mildy- or moderately-depressed subjects. The our analysis,
the binary classification task is \code{depress v control}.

% The intergroup study involved 21 healthy volunteers received a compensation
% for a participation and 51 participants featuring mild depression (F32.0),
% moderate depression (F32.1), or dysthymia (F34.1, single patient) (see Table
% 1). The fMRI study was carried out in the International Tomography Center,
% Novosibirsk, using a 3 T Ingenia scanner (Philips). Functional T2∗-weighted
% Ssh echo planar imaging scans were acquired using the following parameters:
% voxel size 2×2×5mm, repetition time/echo time = 2500/35 ms, and fat
% suppression mode. The reference anatomical image was obtained by the T1W 3D
% turbo field echo method with a voxel size of 1×1×1mm. The instruction for
% participants was to lie still with eyes closed for 6 minutes.

% \subsubsection{Learning Dataset}

The \textit{Learning} dataset\cite{schapiroHumanHippocampalReplay2018,
schapiroHumanHippocampalReplay2020} has both rs-fMRI and task fMRI scans were
available for all subjects, and \code{task v rest} is chosen as the binary
classification task for our analysis.  The task is complex and difficult to
summarize here adequately, but involved multiple phases where subjects were
presented with various related abstract images, and then later tested for their
memory of certain aspects of the learned images. Thus the classification task
might also be considered learning vs. rest.

% \subsubsection{Osteoarthritis Dataset}
The \textit{Osteo} dataset\cite{tetreaultBrainConnectivityPredicts2016}
includes whiole-brain rs-fMRI scans of healthy, pain-free controls
(\code{nopain} condition), and individuals with knee osteoarthritis.
Osteoarthritic patients were treated for 2 weeks with either placebo
(\code{pain} condition) or duloxetine (\code{duloxetine} condition).
We use the three binary classification tasks \code{nopain v duloxetine},
\code{nopain v pain}, and \code{pain v duloxetine} for our analysis.

% \subsubsection{Parkinson's Dataset}
The \textit{Park} dataset\cite{madhyasthaDynamicConnectivityRest2015} includes
subjects with non-demented Parkinson's Disease (PD) and
healthy controls performed a number of repetitions of the Attention Network
Task (ANT; \cite{fanActivationAttentionalNetworks2005}) during scans.
The classification task for this dataset is Parkinson's versus controls, i.e.
\code{park v ctrl}.

% \subsubsection{Attention Dataset}

The \textit{Attention} dataset\cite{gorgolewskiHighResolution7Tesla2015} is a
high-resolution rs-fMRI dataset including a battery of psychological measures
for each subject. Since previous studies employing RMT sometimes interpreted
their findings with respect to attentional processes
\cite{wangRandomMatrixTheory2016,matharooSpontaneousBackpainAlters2020}, we
divided subjects into various high vs. low attention binary classification
tasks based on median splits of subsets of the metrics available in this study.

The \code{viglant v nonvigilant} task was formed based on self-report
questionnaire items involving
"vigilance"\cite{gorgolewskiCorrespondenceIndividualDifferences2014}. The
\code{trait\_attend v trait\_nonattend} task was formed using PANAS-X
\cite{watsonPANASXManualPositive1994} items related to self-reported
wakefulness and attention over the past weeks. Finally, the
\code{task\_attend v tas\_nonattent} classification task was constructed
using the scores of The Conjunctive Continuous Performance Task
(CCPT\cite{shalevConjunctiveContinuousPerformance2011}). This is a behavioural
performance task which requires the subject to quickly and selectively respond
to only certain visual stimuli. Details allowing exact reproduction of these
splits are available in this paper's
\href{https://github.com/DM-Berger/random-matrix-fmri/blob/master/code/rmt/preprocess/unify\_attention\_data.py}{source
code}.

\section{Methods}
\label{sec:methods}

\subsection{fMRI Preprocessing}
\label{sec:preproc}

We limited preprocessing to a simple pipeline of, in order, (1) brain
extraction, (2) slice timing correction, (3), motion correction, and (4),
registration to the MNI152
\cite{fonovUnbiasedAverageAgeappropriate2011,fonovUnbiasedNonlinearAverage2009}
2mm isotropic, asymmetric template version C made available through
\href{https://www.templateflow.org/}{TemplateFlow}
\cite{ciricTemplateFlowFAIRsharingMultiscale2021}.

We limited preprocessing methods to these steps sometimes because data required
for other typical preprocessing steps (e.g. field intensity maps, physiological
measurements), but also because we saved all pipelined intermediates to later
compare the effect of increasing degrees of pre-processing (see Section
\ref{sec:multiverse}), and it was important to limit the number of
preprocessing degrees to prevent excessive computational costs.

Brain extraction was performed first with BET
\cite{smithFastRobustAutomated2002}, and then with with
ANTs\cite{avantsReproducibleEvaluationANTs2011} to clean up residue left behind
by BET. BET results were visually inspected for each image in each dataset, and
BET fractional anisotropy values were modified to ensure brain extraction was
acceptable. Motion-correction was performed with FSL's MCFLIRT
\cite{jenkinsonImprovedOptimizationRobust2002}, and slicetime correction also
with FSL's \code{slicetimer}\cite{jenkinsonFSL2012} tool. Finally, functional
images were registered directly to the MNI152 template via
ANTs\cite{avantsReproducibleEvaluationANTs2011}.

\subsection{Feature Extraction}

For each preprocessing degree and fMRI image, features were extracted from all
\(N\) non-constant brain voxels, where \(N\) may differ for each image. After
this process, each scan yields an \(N \times T\) matrix \(\mathbf{M}\), where
\(T\) is the number of volumes acquired. All features summarize this matrix
of brain voxels \(\mathbf{M}\) in some manner.

\subsubsection{Raw Eigenvalues}
\label{sec:raw-eigs}

The time and space complexities of computing the full \(N \times N\)
correlation matrix of \(\mathbf{M}\), and the subsequent eigenvalues, are too
large to be tractable. However, since transposition does not change the
eigenvalues, we can use the transpose to efficiently compute these eigenvalues
(see Appendix \ref{sec:transpose}). These raw eigenvalues (feature \code{eigs}
in Table \ref{tab:features}), most directly test the idea that eigenvalues
alone are useful functional connectivity features.

However, not all eigenvalues may have predictive utility. We also examine as
features the central eigenvalues (defined as the middle 10\%, 20\%, or 40\% of
the sorted eigenvalues, \code{eigsmiddle} in Table \ref{tab:features}), and
tail eigenvalues (defined as the first and last 5\%,  10\%, and 20\% of each
tail of the spectrum, \code{eigsminmax} in Table \ref{tab:features}).

\subsubsection{RMT Features}
\label{sec:rmt-features}

To compare the observed eigenvalue distribution to some RMT theoretical
predictions, the eigenvalues must be unfolded
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. The
details and motivation behind the unfolding procedure are well documented in
\cite{guhrRandommatrixTheoriesQuantum1998a}. Practically, the unfolding process
can be viewed as a smoothing and rescaling procedure where the originally
observed spectrum \(\lambda_1 \le \dots \le \lambda_n\) is smoothly mapped to
an unfolded spectrum \(e_1, \dots, e_n\), such that \(\sum d_i/n \approx 1\),
if \(d_i = e_{i+1} - e_i \). For most empirical data, the true smoothing
function is not known, and so must be approximated
\cite{guhrRandommatrixTheoriesQuantum1998a,mehtaRandomMatrices2004}. Typically,
a polynomial is chosen \cite{abul-magdUnfoldingSpectrumChaotic2014}.

Given empirically observed eigenvalues \(\Lambda\), the spectral rigidity,
\(\Delta_3(L)\), is calculated for any positive real value \(L <
\max(\Lambda)\) as:
\begin{equation}
\label{eq:rigidity}
\Delta_3(L) = \left \langle \min_{A,B} \frac{1}{L} \int_c^{c+L} \left(  \eta(\lambda) -A \lambda - B \right)^2 \right \rangle_c
\end{equation}
where \(\eta(\lambda)\) is the number of unfolded eigenvalues less than or
equal to \(\lambda\), \(\langle \cdot \rangle_c\) denotes the average with
respect to all starting points \(c\), and where \(A\) and \(B\) denote the
slope and intercept, respectively, of the least squares fit of a straight line
to \(\eta(\lambda)\) on \([c, c+L]\)
\cite{guhrRandommatrixTheoriesQuantum1998a}. Viewing the unfolded eigenvalues
as a timeseries, the spectral rigidty for a value \(L\) can be is thus the
\textit{average nonlinearity} of all intervals of length \(L\) over the series.

>>>>

% \documentclass[12pt]{supplementary}  % 12pt font required by SPIE;
\documentclass{article}  % use this instead for A4 paper
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{authblk}
\usepackage{times}
\usepackage{float}
\usepackage{ifthen}
\usepackage{setspace}
\usepackage[super]{cite}[2003/11/04] % need vers. > 4.01
\usepackage{color}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{graphicx}
\graphicspath{ {./figures/} }
\usepackage{setspace}
\usepackage{tocloft}
\usepackage{doi}
\usepackage{csquotes}
\MakeOuterQuote{"}
\usepackage{pdflscape}
\usepackage{geometry}
\geometry{
  textwidth=11in,
  textheight=8.5in,
  headheight=0in,
  headsep=0in,
  marginparwidth=0in,
  papersize={11in,8.5in},
  left=0in,
  lmargin=0in,
  inner=0in,
  right=0in,
  rmargin=0in,
  outer=0in,
  top=0in,
  tmarg=0in,
  bottom=0in,
  bmargin=0in,
}

@@

\pagestyle{empty}               % page numbers is default; use empty for no numbers

\title{Supplementary Material}

@@

\renewcommand{\cftdotsep}{\cftnodots}
\cftpagenumbersoff{figure}
\cftpagenumbersoff{table}

\begin{document}

\begin{center}
\maketitle
\end{center}

\begin{spacing}{1}   % use single spacing for supplementary

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth,height=0.9\textheight,keepaspectratio]{all_by_fine_feature_groups.png}
\end{center}
\caption
{ \label{fig:feature-group-all}
AUROC distributions across gross feature groupings and comparison tasks.}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth,height=0.9\textheight,keepaspectratio]{fine_feature_by_preproc_predictive_subgroup.png}
\end{center}
\caption
{ \label{fig:fine-preproc}
AUROC distributions across fine feature groupings and predictable comparison tasks, with effect of preprocessing.}
\end{figure}
